Scientific Reasoning: \\ Assessment of Multimodal Generative LLMs

### 2024-12-10 docs/paper/main.tex
corrections

\paragraph{Cosine similarity with Sentence Transformers}
Cosine similarity using SentenceTransformer embeddings provides a robust method for evaluating model outputs in the ScienceQA dataset, focusing on the semantic alignment of generated explanations with references. It captures nuanced relationships and conceptual understanding, which gives more understanding of the generated outputs in comparison to ROUGE and BLEU scores and that is why it has more importance in final score computation. However, due to domain specificity it can miss some intricate cases like still considering high similarity of two sentences, even though it the reference can contain crucial information about carbon dioxide, while in the candidate sentence the answer was given in relation to oxygen reproduction.
\paragraph{METEOR}
METEOR (\cite{banerjee2005meteor}) evaluates text similarity based on precision, recall, and alignment of semantically meaningful components like synonyms and stems, making it well-suited for capturing the nuanced language used in scientific reasoning.
By incorporating a weighted F-score and penalties for excessive mismatches, METEOR provides a balanced evaluation of fluency and accuracy, crucial for assessing LLM-generated answers on the Science QA Dataset. Due to these balanced text evaluation, this metric has \textbf{10\%} impact on the overall score.
% \subsubsection{Cosine similarity with SciBERT}
% We extended the evaluation of cosine similarity using SciBERT (\cite{beltagy2019scibert}) embeddings. These vectors are more consistent with the domain-specific terminology and reasoning patterns present in the Science QA Dataset because it has been properly pre-trained on scientific literature.
% SciBERT generates more accurate similarity embeddings because it better reflects the complex relationships between scientific concepts than general-purpose models like those in Sentence Transformers.
% When assessing the logical and factual coherence of scientific reasoning, the structural and contextual semantics specific to scientific texts are reflected in the embeddings produced by SciBERT.
% Assessments for jobs requiring sophisticated scientific understanding are more reliable when SciBERT is used for cosine similarity since it guarantees that the evaluation metric is domain-specific.

\paragraph{Cosine similarity with Sentence Transformers}
Sentence Transformer models are optimized for sentence-level embeddings, which capture the overall contextual similarity of answers more effectively, making them suitable for general-purpose evaluation tasks like scientific reasoning.
Unlike SciBERT, which is tailored to scientific text and could have been used as sentence embedder, Sentence Transformers are pre-trained on diverse datasets, enabling them to generalize better to variations in phrasing and style present in the Science QA Dataset.
Their ability to handle broader linguistic nuances ensures that the evaluation accounts for both semantic correctness and the logical flow of responses, which is essential when assessing reasoning across a wide range of scientific topics. That is the reason for using Sentence Transformer model for measurements of cosine similarity between the lecture material and annotated explanations and models' outputs with impact weight of \textbf{15\%}.

\paragraph{Overall score}
In order to obtain a numeric description of the model's performance, we set the follwoing metric that includes previously mentioned metrics with the given importance weights:
\begin{equation*}
\begin{split}
\text{Overall Score} = & \, 0.6 \cdot \text{Accuracy} + 0.05 \cdot \text{BLEU-1} \\
& + 0.5 \cdot \text{BLEU-4} + 0.05 \cdot \text{ROUGE} \\
& + 0.15 \cdot \text{METEOR} \\
& + 0.1 \cdot \text{Cosine Similarity}
\end{split}
\end{equation*}

>>>>

\paragraph{Cosine similarity with Sentence Transformers}
Cosine similarity using SentenceTransformer embeddings provides a robust method for evaluating model outputs in the ScienceQA dataset, focusing on the semantic alignment of generated explanations with references. It captures nuanced relationships and conceptual understanding, which gives more understanding of the generated outputs in comparison to ROUGE and BLEU scores and that is why it has more importance in final score computation. However, due to domain specificity it can miss some intricate cases like still considering high similarity of two sentences, even though it the reference can contain crucial information about carbon dioxide, while in the candidate sentence the answer was given in relation to oxygen reproduction.
\paragraph{METEOR}
METEOR (\cite{banerjee2005meteor}) evaluates text similarity based on precision, recall, and alignment of semantically meaningful components like synonyms and stems, making it well-suited for capturing the nuanced language used in scientific reasoning.
By incorporating a weighted F-score and penalties for excessive mismatches, METEOR provides a balanced evaluation of fluency and accuracy, crucial for assessing LLM-generated answers on the Science QA Dataset. Due to these balanced text evaluation, this metric has \textbf{15\%} impact on the overall score.
% \subsubsection{Cosine similarity with SciBERT}
% We extended the evaluation of cosine similarity using SciBERT (\cite{beltagy2019scibert}) embeddings. These vectors are more consistent with the domain-specific terminology and reasoning patterns present in the Science QA Dataset because it has been properly pre-trained on scientific literature.
% SciBERT generates more accurate similarity embeddings because it better reflects the complex relationships between scientific concepts than general-purpose models like those in Sentence Transformers.
% When assessing the logical and factual coherence of scientific reasoning, the structural and contextual semantics specific to scientific texts are reflected in the embeddings produced by SciBERT.
% Assessments for jobs requiring sophisticated scientific understanding are more reliable when SciBERT is used for cosine similarity since it guarantees that the evaluation metric is domain-specific.

\paragraph{Cosine similarity with Sentence Transformers}
Sentence Transformer models are optimized for sentence-level embeddings, which capture the overall contextual similarity of answers more effectively, making them suitable for general-purpose evaluation tasks like scientific reasoning.
Unlike SciBERT, which is tailored to scientific text and could have been used as sentence embedder, Sentence Transformers are pre-trained on diverse datasets, enabling them to generalize better to variations in phrasing and style present in the Science QA Dataset.
Their ability to handle broader linguistic nuances ensures that the evaluation accounts for both semantic correctness and the logical flow of responses, which is essential when assessing reasoning across a wide range of scientific topics. That is the reason for using Sentence Transformer model for measurements of cosine similarity between the lecture material and annotated explanations and models' outputs with impact weight of \textbf{10\%}.

\paragraph{Overall score}
In order to obtain a numeric description of the model's performance, we set the follwoing metric that includes previously mentioned metrics with the given importance weights:
\begin{equation*}
\begin{split}
\text{Overall Score} = & \, 0.6 \cdot \text{Accuracy} + 0.05 \cdot \text{BLEU-1} \\
& + 0.5 \cdot \text{BLEU-4} + 0.05 \cdot \text{ROUGE} \\
& + 0.15 \cdot \text{METEOR} \\
& + 0.1 \cdot \text{Cosine Similarity}
\end{split}
\end{equation*}
We prioritize the ability to generate correct answers as the most critical aspect, with reasoning ranked as the second-highest priority in our overall evaluation metric.
